EldersAIR initial Micropem data work

### 2020-09-02 micropem_data_cleaning.Rmd
micropem cleaning updates

elders_ids_linked <- read_rds(paste0(file_path, "Output/elders_ids_linked.rds")) %>% 
  filter(winter_id == 1 | winter_id == 2)

@@

# Format combined file from above; combine with other PM data below
np_micropem <- initial_data %>% 
  separate(time, c("trash1", "time_pm"), sep = " ", remove = FALSE) %>% 
  separate(list_files, c("file_name", "trash2"), sep = ".xlsx", remove = FALSE) %>%
  mutate(date_pm = ymd(date),
         pm_personal = as.numeric(pm),
         area = "NPT",
         file_name = as.character(file_name),
         file_type = "xlsx") %>% 
  unite("datetime", c("date_pm", "time_pm"), sep = " ", remove = FALSE) %>% 
  mutate(datetime_pm = ymd_hms(datetime)) %>% 
  select(file_name, area, datetime_pm, date_pm, time_pm, pm_personal, file_type,
         temp:shutdown_reason) %>% 
  filter(!is.na(pm_personal)) %>% 
  ungroup() %>% 
  arrange(area, file_name, datetime_pm)

# save files
#write_rds(np_micropem, paste0(file_path, "Output/np_micropem.rds"))

### NN ###
# List files
list_files <- list.files(paste0(file_path, "Input/NN/nn_micropem_small_20200302"))

# Set working directory and load files in list; extract file name and add as column
## run next lines together
setwd(paste0(file_path, "Input/NN/nn_micropem_small_20200302")) # set new working directory
initial_data <- tibble(list_files) %>% # use list of files from above
  mutate(save_data = lapply(list_files, read_csv, skip = 28,
                            col_names = c("date", "time", "pm", "pm2",
                                          "temp", "rh", "battery",
                                          "in_pressure", "or_pressure",
                                          "flow", "x_axis", "y_axis",
                                          "z_axis", "vector_sum", "shutdown_reason",
                                          "trash1", "trash2"),
                            col_types = list(col_character(), col_character(), col_double(),
                                          col_double(), col_double(), col_double(),
                                          col_double(), col_double(), col_double(),
                                          col_double(), col_double(), col_double(),
                                          col_double(), col_double(), col_character(),
                                          col_character(), col_character()))) %>%
  unnest(save_data)

list_files <- list.files(paste0(file_path, "Input/NN/nn_micropem_large_20200302"))

# Set working directory and load files in list; extract file name and add as column
## run next lines together
setwd(paste0(file_path, "Input/NN/nn_micropem_large_20200302")) # set new working directory
initial_data2 <- tibble(list_files) %>% # use list of files from above
  mutate(save_data = lapply(list_files, read_csv, skip = 28,
                            col_names = c("date", "time", "pm", "pm2",
                                          "temp", "rh", "battery",
                                          "in_pressure", "or_pressure",
                                          "flow", "x_axis", "y_axis",
                                          "z_axis", "vector_sum", "shutdown_reason",
                                          "trash1", "trash2"),
                            col_types = list(col_character(), col_character(), col_double(),
                                          col_double(), col_double(), col_double(),
                                          col_double(), col_double(), col_double(),
                                          col_double(), col_double(), col_double(),
                                          col_double(), col_double(), col_character(),
                                          col_character(), col_character()))) %>%
  unnest(save_data)

nn_micropem_raw <- rbind(initial_data, initial_data2)

# Format combined file from above; combine with other PM data below
nn_micropem <- nn_micropem_raw %>% 
  filter(!is.na(pm)) %>% 
  separate(list_files, c("file_name", "trash2"), sep = ".csv", remove = FALSE) %>%
  mutate(date_pm = mdy(date),
         time_pm = time,
         pm_personal = as.numeric(pm),
         area = "NN",
         file_name = as.character(file_name),
         file_type = "csv") %>% 
  unite("datetime", c("date_pm", "time_pm"), sep = " ", remove = FALSE) %>% 
  mutate(datetime_pm = ymd_hms(datetime)) %>% 
  select(file_name, area, datetime_pm, date_pm, time_pm, pm_personal, file_type,
         temp:shutdown_reason) %>% 
  ungroup() %>% 
  arrange(area, file_name, datetime_pm)

# Save file
#write_rds(nn_micropem, paste0(file_path, "Output/nn_micropem.rds"))

# Bind files from both areas
elders_micropem <- rbind(nn_micropem, np_micropem)

# save joined file
#write_rds(elders_micropem, paste0(file_path, "Output/elders_micropem.rds"))
```

@@

elders_micropem_log <- read_rds(paste0(file_path, "Output/elders_micropem_log.rds")) %>% 
  select(home, adult_id_num, adult_id_char, home_winter_id:winter_id, 
         -sample_obs, treatment)

np_micropem <- read_rds(paste0(file_path, "Output/np_micropem.rds")) %>% 
  rename(mp_file_name = file_name)

elders_micropem_new <- np_micropem %>% 
  left_join(elders_micropem_log, by = c("area", "mp_file_name")) %>% 
  mutate(datetime_pm = ymd_hms(datetime_pm),
         start_datetime_pm = ymd_hms(start_datetime_pm),
         stop_datetime_pm = ymd_hms(stop_datetime_pm))

nn_micropem_new <- elders_micropem_new %>% 
  filter(area == "NN") %>% 
  mutate(date_filter = if_else(datetime_pm >= start_datetime_pm & 
                               datetime_pm <= stop_datetime_pm, 1, 0))

nn_micropem_new2 <- nn_micropem_new %>% 
  filter(date_filter == 1)

check_obs <- nn_micropem_new2 %>% 
  count(home_winter_id)

# Save data
#write_rds(nn_micropem_new2, paste0(file_path, "Output/nn_micropem_new.rds"))

np_micropem_new <- elders_micropem_new %>% 
  filter(area == "NPT") %>% 
  mutate(date_filter = if_else(datetime_pm >= start_datetime_pm & 
                               datetime_pm <= stop_datetime_pm, 1, 0))

np_micropem_new2 <- np_micropem_new %>% 
  filter(date_filter == 1)

# Save data
#write_rds(np_micropem_new2, paste0(file_path, "Output/np_micropem_new.rds"))

elders_micropem_new <- rbind(nn_micropem_new2, np_micropem_new2)

# Save data
#write_rds(elders_micropem_new, paste0(file_path, "Output/elders_micropem_new.rds"))
```

# Summarize filtered micropem data for further cleaning
```{r}
elders_micropem <- read_rds(paste0(file_path, "Output/elders_micropem_new.rds"))

micropem_observations_summary <- elders_micropem %>% 
  group_by(area, home, adult_id_char, winter_id, sampling_visit) %>% 
  count(file_name)
head(micropem_observations_summary, 10)

micropem_pm_summary <- elders_micropem %>% 
  #group_by(area, winter_id, sampling_visit) %>%
  group_by(area, file_name) %>% 
  summarize("n" = n(),
            "Mean PM" = mean(pm_personal, na.rm = TRUE), 
            "SD PM" = sd(pm_personal, na.rm = TRUE),
            "Min PM" = min(pm_personal, na.rm = TRUE), 
            "25%" = quantile(pm_personal, 0.25, na.rm = TRUE),
            "Median PM" = median(pm_personal, na.rm = TRUE),
            "75%" = quantile(pm_personal, 0.75, na.rm = TRUE),
            "Max PM" = max(pm_personal, na.rm = TRUE))
micropem_pm_summary
```

# Continue micropem cleaning
```{r}
elders_micropem <- read_rds(paste0(file_path, "Output/elders_micropem_new.rds"))

elders_micropem_new2 <- elders_micropem %>%   
  group_by(area, file_name) %>% 
  # new variables summarizing pm2.5 within each home
  mutate(pm_total_observations = n(),
         # new datetime values to calculate sampling interval
         first_datetime = first(datetime_pm),
         last_datetime = last(datetime_pm),
         first_datetime = ymd_hms(first_datetime),
         last_datetime = ymd_hms(last_datetime),
         diff_datetime = interval(first_datetime, last_datetime),
         # total sampling interval, based on first and last datetimes
         pm_sample_interval = diff_datetime/86400,
         # expected number of observations with the sampling interval
         expected_obs = pm_sample_interval*24*60,
         # difference between actual and expected observations
         actual_vs_expected_diff = pm_total_observations - expected_obs,
         # indicator variable if actual vs expected obs > 10 or greater
         actual_vs_expected = if_else(actual_vs_expected_diff < 10 &
                                      actual_vs_expected_diff > -10, 1, 0),
         pm_mean = mean(pm_personal),
         # breaking sampling period into days
         obs = row_number(),
         sampling_day = if_else(obs < 1441, 1, 0),
         sampling_day = if_else(obs < 2881 & obs > 1440, 2, sampling_day),
         sampling_day = if_else(obs < 4321 & obs > 2880, 3, sampling_day)) %>%
  # calculate pm means for each sampling day
  group_by(area, file_name, sampling_day) %>% 
  mutate(pm_mean_daily = mean(pm_personal),
         pm_sampling_day = sampling_day,
         day_of_week = weekdays(date_pm)) %>%
  mutate(start_datetime_pm = ymd_hms(start_datetime_pm),
         pm_start_diff = interval(first_datetime, start_datetime_pm),
         pm_start_diff_hrs = pm_start_diff/3600) %>% 
  ungroup() %>% 
  select(-obs, -sampling_day) %>% 
  arrange(area, home, sampling_visit) 
# write_rds(elders_micropem_new2, paste0(file_path, "Output/elders_micropem_new2.rds"))
```

# Function to look at PM data across time
```{r}
# Plot PM by datetime to look at trends, gaps, and peaks
pm_time_plot_function <- function(id, location) {

pm_time_plot <- elders_micropem %>% 
  filter(home_winter_id == id & area == location) %>% 
  filter(winter_id == 1) %>% 
  arrange(datetime_pm) %>% 
  mutate(pm_sampling_day = as.character(pm_sampling_day),
         day_of_week = factor(day_of_week,
                              levels = c("Sunday", "Monday", "Tuesday", "Wednesday", 
                                         "Thursday", "Friday", "Saturday"))) %>%
  ggplot() + 
    geom_point(aes(datetime_pm, pm_personal, color = day_of_week), size = 1.5) +
    theme_classic() +
    labs(title = id,
         y = expression(paste("PM"[2.5], " (", mu, g/m^3, ")")),
         x = "Datetime",
         color = "Day of week") +
    theme(axis.title.y = element_text(size = 12,
                                      margin = margin(t = 0, r = 20, b = 0, l = 0)),
          axis.title.x = element_text(size = 12),
          axis.text.x = element_blank(),
          axis.text.y = element_text(size = 12, color = "black"),
          axis.line.x = element_line(colour = "black", size = 1), 
          axis.line.y = element_line(colour = "black", size = 1), 
          axis.ticks = element_blank()) +
    scale_color_manual(values = jv_palette)

gaps <<- elders_micropem %>% 
  filter(home_winter_id == id & area == location) %>% 
  mutate(datetime_pm = ymd_hms(datetime_pm),
         datetime_lead = ymd_hms(lead(datetime_pm)),
         datetime_1 = seconds(datetime_pm),
         datetime_2 = seconds(datetime_lead),
         datetime_diff_minutes = as.numeric(datetime_2 - datetime_1)/60,
         datetime_diff_hours = as.numeric(datetime_diff_minutes/60)) %>% 
  arrange(desc(datetime_diff_minutes)) %>% 
  select(day_of_week, pm_sampling_day, datetime_diff_minutes, datetime_diff_hours, 
         datetime_pm, datetime_lead)

pm_desc <<- elders_micropem %>% 
  filter(home_winter_id == id & area == location) %>% 
  arrange(desc(pm_personal)) %>% 
  select(datetime_pm, pm_personal, pm_sampling_day)
pm_time_plot

@@

# Run function from above to look at results
## Function run for each household to check for odd trends/gaps in the data
## Notes from this process on uploaded to Box:
### EldersAIR > Ethan - data cleaning and analysis > pm_cleaning_summary
```{r}
# load data
elders_micropem <- read_rds(paste0(file_path, "Output/elders_micropem_new2.rds"))

# Change function input to look at specific home
# Run all of these lines together
pm_time_plot_function(id = "37", location = "NN")
head(gaps, 25)
head(pm_desc, 25)
tail(pm_desc, 25)
# filter data for a single home
check_data <- elders_micropem %>% 
  filter(area == "NN" & home_winter_id == "37") %>% 
  filter(pm_personal < 0)
```

>>>>

elders_ids_linked <- read_rds(paste0(file_path, "Output/elders_ids_linked.rds")) %>% 
  filter(winter_id == 1 | winter_id == 2) %>% 
  select(-adult_id_num, -adult_id_char) %>% 
  distinct(area, home_winter_id, .keep_all = TRUE)

@@

# Format combined file from above; combine with other PM data below
npt_micropem <- initial_data %>% 
  separate(time, c("trash1", "time_pm"), sep = " ", remove = FALSE) %>% 
  separate(list_files, c("mp_file_name", "trash2"), sep = ".xlsx", remove = FALSE) %>%
  mutate(date_pm = ymd(date),
         pm_personal = as.numeric(pm),
         area = "NPT",
         mp_file_name = as.character(mp_file_name),
         file_type = "xlsx") %>% 
  unite("datetime", c("date_pm", "time_pm"), sep = " ", remove = FALSE) %>% 
  mutate(datetime_pm = ymd_hms(datetime)) %>% 
  select(mp_file_name, area, datetime_pm, date_pm, time_pm, pm_personal, file_type,
         temp:shutdown_reason) %>% 
  filter(!is.na(pm_personal)) %>% 
  ungroup() %>% 
  arrange(area, mp_file_name, datetime_pm)

# save files
write_rds(npt_micropem, paste0(file_path, "Output/npt_micropem.rds"))

### NN ###
# Only doing NPT micropem data for now, as NN data is a mess and unreliable
```

@@

npt_micropem_log <- read_rds(paste0(file_path, "Output/elders_micropem_log.rds")) %>% 
  filter(area == "NPT") %>% 
  filter(!is.na(stop_date)) %>% 
  select(area, home, home_winter_id:winter_id, 
         -sample_obs, treatment)

npt_micropem <- read_rds(paste0(file_path, "Output/npt_micropem.rds"))

npt_micropem_new <- npt_micropem %>% 
  left_join(npt_micropem_log, by = c("area", "mp_file_name")) %>% 
  mutate(datetime_pm = ymd_hms(datetime_pm),
         start_datetime_pm = ymd_hms(start_datetime_pm),
         stop_datetime_pm = ymd_hms(stop_datetime_pm))

# Filter to only dates that fall within sampling timeframe
npt_micropem_new2 <- npt_micropem_new %>% 
  mutate(date_filter = if_else(datetime_pm >= start_datetime_pm & 
                               datetime_pm <= stop_datetime_pm, 1, 0))

npt_micropem_new3 <- npt_micropem_new2 %>% 
  filter(date_filter == 1)

# Save data
write_rds(npt_micropem_new3, paste0(file_path, "Output/npt_micropem_new.rds"))

check <- npt_micropem_log %>% 
  distinct(mp_file_name, .keep_all = TRUE)
```

# Summarize filtered micropem data for further cleaning
```{r}
file_path <- c("C:/Users/ethan.walker/Box/Ethan Walker UM/R/EldersAIR/")

npt_micropem_new <- read_rds(paste0(file_path, "Output/npt_micropem_new.rds"))

micropem_observations_summary <- npt_micropem_new %>% 
  group_by(area, home, winter_id, sampling_visit) %>% 
  count(mp_file_name)
head(micropem_observations_summary, 10)

micropem_pm_summary <- npt_micropem_new %>% 
  #group_by(area, winter_id, sampling_visit) %>%
  group_by(area, mp_file_name) %>% 
  summarize("n" = n(),
            "Mean PM" = mean(pm_personal, na.rm = TRUE), 
            "SD PM" = sd(pm_personal, na.rm = TRUE),
            "Min PM" = min(pm_personal, na.rm = TRUE), 
            "25%" = quantile(pm_personal, 0.25, na.rm = TRUE),
            "Median PM" = median(pm_personal, na.rm = TRUE),
            "75%" = quantile(pm_personal, 0.75, na.rm = TRUE),
            "Max PM" = max(pm_personal, na.rm = TRUE))
micropem_pm_summary
```

# Continue micropem cleaning
```{r}
file_path <- c("C:/Users/ethan.walker/Box/Ethan Walker UM/R/EldersAIR/")

npt_micropem <- read_rds(paste0(file_path, "Output/npt_micropem_new.rds"))

npt_micropem_new <- npt_micropem %>%   
  group_by(area, mp_file_name) %>% 
  # new variables summarizing pm2.5 within each home
  mutate(pm_total_observations = n(),
         # new datetime values to calculate sampling interval
         first_datetime = first(datetime_pm),
         last_datetime = last(datetime_pm),
         first_datetime = ymd_hms(first_datetime),
         last_datetime = ymd_hms(last_datetime),
         diff_datetime = interval(first_datetime, last_datetime),
         # total sampling interval, based on first and last datetimes
         pm_sample_interval = diff_datetime/86400,
         # expected number of observations with the sampling interval
         expected_obs = pm_sample_interval*24*60,
         # difference between actual and expected observations
         actual_vs_expected_diff = pm_total_observations - expected_obs,
         # indicator variable if actual vs expected obs > 10 or greater
         actual_vs_expected = if_else(actual_vs_expected_diff < 10 &
                                      actual_vs_expected_diff > -10, 1, 0),
         pm_mean = mean(pm_personal, na.rm = TRUE),
         pm_sd = sd(pm_personal, na.rm = TRUE),
         pm_min = min(pm_personal, na.rm = TRUE),
         pm_median = median(pm_personal, na.rm = TRUE),
         pm_max = max(pm_personal, na.rm = TRUE),
         # breaking sampling period into days
         obs = row_number(),
         sampling_day = if_else(obs < 1441, 1, 0),
         sampling_day = if_else(obs < 2881 & obs > 1440, 2, sampling_day),
         sampling_day = if_else(obs < 4321 & obs > 2880, 3, sampling_day)) %>%
  # calculate pm means for each sampling day
  group_by(area, mp_file_name, sampling_day) %>% 
  mutate(pm_mean_daily = mean(pm_personal, na.rm = TRUE),
         pm_sampling_day = sampling_day,
         day_of_week = weekdays(date_pm)) %>%
  mutate(start_datetime_pm = ymd_hms(start_datetime_pm),
         pm_start_diff = interval(first_datetime, start_datetime_pm),
         pm_start_diff_hrs = pm_start_diff/3600) %>% 
  ungroup() %>% 
  select(-obs, -sampling_day) %>% 
  arrange(area, home, sampling_visit)

write_rds(npt_micropem_new, paste0(file_path, "Output/npt_micropem_new2.rds"))

pm_summary <- npt_micropem_new %>% 
  distinct(mp_file_name, .keep_all = TRUE) %>% 
  select(-datetime_pm, -date_pm, -time_pm, -pm_personal, -file_type,
         -x_axis, -y_axis, -z_axis, -vector_sum,
         -start_date, -start_time, -stop_date, -stop_time, -day_of_week,
         -fb_file_name, -treatment, -date_filter, -pm_mean_daily, -pm_sampling_day) %>% 
  select(mp_file_name:sampling_visit, winter_id, start_datetime_pm,
         first_datetime, stop_datetime_pm, last_datetime, diff_datetime,
         pm_sample_interval, pm_total_observations, expected_obs, actual_vs_expected_diff,
         actual_vs_expected:pm_start_diff_hrs)
write_csv(pm_summary, paste0(file_path, "Output/npt_pm_summary.csv"))
```

# Function to look at PM data across time
```{r}
# Plot PM by datetime to look at trends, gaps, and peaks
jv_palette <- c("#330099","#CC0066","#FF6633", 
                 "#0099CC", "#FF9900","#CC6633",
                  "#FF3366", "#33CC99", "#33999")

pm_time_plot_function <- function(id, location, visit) {

pm_time_plot <- npt_micropem %>% 
  filter(home_winter_id == id & area == location & sampling_visit == visit) %>% 
  #filter(pm < 30000) %>% 
  arrange(datetime_pm) %>% 
  mutate(pm_sampling_day = as.character(pm_sampling_day),
         day_of_week = factor(day_of_week,
                              levels = c("Sunday", "Monday", "Tuesday", "Wednesday", 
                                         "Thursday", "Friday", "Saturday"))) %>%
  ggplot() + 
    geom_point(aes(datetime_pm, pm_personal, color = day_of_week), size = 1.5) +
    theme_classic() +
    labs(title = id,
         y = expression(paste("PM"[2.5], " (", mu, g/m^3, ")")),
         x = "Datetime",
         color = "Day of week") +
    theme(axis.title.y = element_text(size = 12,
                                      margin = margin(t = 0, r = 20, b = 0, l = 0)),
          axis.title.x = element_text(size = 12),
          axis.text.x = element_blank(),
          axis.text.y = element_text(size = 12, color = "black"),
          axis.line.x = element_line(colour = "black", size = 1), 
          axis.line.y = element_line(colour = "black", size = 1), 
          axis.ticks = element_blank()) +
    scale_color_manual(values = jv_palette)

gaps <<- npt_micropem %>% 
  filter(home_winter_id == id & area == location & sampling_visit == visit) %>% 
  mutate(datetime_pm = ymd_hms(datetime_pm),
         datetime_lead = ymd_hms(lead(datetime_pm)),
         datetime_1 = seconds(datetime_pm),
         datetime_2 = seconds(datetime_lead),
         datetime_diff_minutes = as.numeric(datetime_2 - datetime_1)/60,
         datetime_diff_hours = as.numeric(datetime_diff_minutes/60)) %>% 
  arrange(desc(datetime_diff_minutes)) %>% 
  select(day_of_week, pm_sampling_day, datetime_diff_minutes, datetime_diff_hours, 
         datetime_pm, datetime_lead)

pm_desc <<- npt_micropem %>% 
  filter(home_winter_id == id & area == location & sampling_visit == visit) %>% 
  arrange(desc(pm_personal)) %>% 
  select(datetime_pm, pm_personal, pm_sampling_day)
pm_time_plot

@@

# Run function from above to look at results
## Function run for each household to check for odd trends/gaps in the data
## Notes from this process on uploaded to Box:
### EldersAIR > Ethan - data cleaning and analysis > pm_cleaning_summary
```{r}
# load data
file_path <- c("C:/Users/ethan.walker/Box/Ethan Walker UM/R/EldersAIR/")

npt_micropem <- read_rds(paste0(file_path, "Output/npt_micropem_new2.rds"))

# Change function input to look at specific home
# Run all of these lines together
pm_time_plot_function(id = "51", location = "NPT", visit = 2)
head(gaps, 10)
head(pm_desc, 10)
tail(pm_desc, 10)
# filter data for a single home
check_data <- elders_micropem %>% 
  filter(area == "NPT" & home_winter_id == "37" & sampling_visit == 1) %>% 
  filter(pm_personal < 0)
```
